Characteristics of Top Songs Has Changed from Pandemic Brain

### 2024-04-18 paper/paper.qmd
Updated abstract, introduction, and model results

---
title: "Characteristics of Top Songs Has Changed from Pandemic Brain"
subtitle: "An analysis of songs on Billboard's Year-End Hot 100 singles list (2014 to 2023)"
author: Emily Su
thanks: "Code and data are available at: https://github.com/moonsdust/top-songs."
date: today
date-format: long
abstract: "Music often reflects the current climate of society and there is a growing interest in how hit songs as seen on Billboard's Year-End Hot 100 singles has changed after the COVID-19 pandemic. This paper looks at music characteristics such as tempo, song duration, loudness, and modality of songs from Billboard's Year-End Hot 100 singles list from 2014 to 2023 to reveal patterns and relationships to explain the difference between top songs before 2020 and 2020 onwards. The results show that hit songs from 2020 onwards had become on average shorter, quieter, but slightly faster with the melody of songs being in a major key. These results can support the evaluation of the emotional state of different populations and improve treatments such as music therapy, however further investigation is needed on the influence of lyrics on different music characteristics."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

In this paper, to investigate patterns and trends in music, we analyzed data from the Billboard Year-End Hot 100 singles list from 2014 to 2023 on music characteristics such as its modality (major or minor key), track duration, loudness, and tempo. Our estimand is the song characteristics from the Billboard Year-End Hot 100 singles such as its modality (major or minor), track duration, loudness, and tempo if a song was created before 2020 or during and after 2020. The list allowed us to gauge how preference in music characteristics has changed overall during and after the pandemic. Current studies give us a sense of how currently people are feeling based on surveys done. However, there is currently a lack of understanding and focus on the difference in musical characteristics of popular songs with the general population before the pandemic and during and after it. In our findings, our data showed that songs in a major key made up the majority of top songs before, during, and after 2020 and top songs were on average quieter, shorter, and had slightly faster tempos during and after 2020. [TODO: ADD RESULTS FROM MODEL]. Music can reflect the climate of society and understanding trends in music characteristics of hit songs can help provide insight on the overall feelings of different populations during and after the pandemic as well as before it. This can aid in improving treatments such as music therapy.

@@

@tbl-model-result indicates that the intercept $\beta_0$ is -2.002. This means that songs in a major key with tempo, loudness, and mode being 0 is less likely to be a top song before 2020. @tbl-model-result and @fig-model-result-ci reveal that the coefficient of track duration is 0.000 when tempo, mode, and loudness remains constant and its 90% credible interval includes the value 0. This suggest that when the duration of the song changes the likelihood of a song being a top song prior to 2020 does not change or is impacted. When loudness varies and track duration, tempo, and mode stays fixed, @tbl-model-result indicates that its coefficient is 0.148 with a 90% credible interval above 0 as seen in @fig-model-result-ci. As the coefficient estimate is positive, it means that hit songs that are louder are more likely to be a hit song prior to 2020. Since the coefficient estimate for tempo is −0.002 and its 90% credible interval being slightly below 0 when only tempo varies, it means songs with slower tempos are less likely to be a hit song prior to 2020. The coefficient estimate for the modality/mode is 0.095 and its 90% credible interval being wider suggests that top songs that are in a minor key are likely to be a hit before 2020. However, the true value for the coefficient for mode can also be negative according to @fig-model-result-ci since the interval passes through 0.

>>>>

---
title: "Characteristics of Top Songs Has Changed from Pandemic Brain"
subtitle: "An analysis of songs on Billboard's Year-End Hot 100 singles list (2014 to 2023)"
author: Emily Su
thanks: "Code and data are available at: https://github.com/moonsdust/top-songs."
date: today
date-format: long
abstract: "Music often reflects the current climate of society and there is a growing interest in how hit songs as seen on Billboard's Year-End Hot 100 singles has changed after the COVID-19 pandemic. This paper looks at music characteristics such as tempo, song duration, loudness, and modality of songs from Billboard's Year-End Hot 100 singles list from 2014 to 2023 to reveal patterns and relationships to explain the difference between top songs before 2020 and 2020 onwards. The results show that hit songs from 2020 onwards had become on average quieter but slightly faster with our model revealing that duration does not impact likelihood of a song being a hit prior to 2020 but songs in a minor key were more likely to be a hit before 2020. These results can support the evaluation of the emotional state of different populations and improve treatments such as music therapy, however further investigation is needed on the influence of lyrics on different music characteristics."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

In this paper, to investigate patterns and trends in music, we analyzed data from the Billboard Year-End Hot 100 singles list from 2014 to 2023 on music characteristics such as its modality (major or minor key), track duration, loudness, and tempo. Our estimand is the song characteristics from the Billboard Year-End Hot 100 singles such as its modality (major or minor), track duration, loudness, and tempo if a song was created before 2020 or during and after 2020. The list allowed us to gauge how preference in music characteristics has changed overall during and after the pandemic. Current studies give us a sense of how currently people are feeling based on surveys done. However, there is currently a lack of understanding and focus on the difference in musical characteristics of popular songs with the general population before the pandemic and during and after it. In our findings, our data showed that songs in a major key made up the majority of top songs before, during, and after 2020 and top songs were on average quieter, shorter, and had slightly faster tempos during and after 2020. In our model, it revealed that top songs prior to 2020 are more likely to be louder, slower, and if its in a minor key, its more likely to be a hit song before 2020. However, duration did not impact the likelihood in our model. Music can reflect the climate of society and understanding trends in music characteristics of hit songs can help provide insight on the overall feelings of different populations during and after the pandemic as well as before it. This can aid in improving treatments such as music therapy.

@@

@tbl-model-result indicates that the intercept $\beta_0$ is -2.002. This means that songs in a major key with tempo, loudness, and mode being 0 is less likely to be a top song before 2020. @tbl-model-result and @fig-model-result-ci reveal that the coefficient estimate of track duration $\beta_1$ is 0.000 when tempo, mode, and loudness remains constant and its 90% credible interval includes the value 0. This suggest that when the duration of the song changes the likelihood of a song being a top song prior to 2020 does not change or is impacted. When loudness varies and track duration, tempo, and mode stays fixed, @tbl-model-result indicates that its coefficient estimate for $\beta_2$ is 0.148 with a 90% credible interval above 0 as seen in @fig-model-result-ci. As the coefficient estimate is positive, it means that hit songs that are louder are more likely to be a hit song prior to 2020. Since the coefficient estimate for tempo $\beta_3$ is −0.002 and its 90% credible interval being slightly below 0 when only tempo varies, it means songs with slower tempos are more likely to be a hit song prior to 2020. The coefficient estimate for the modality (mode) $\beta_4$ is 0.095 and its 90% credible interval being wider suggests that top songs that are in a minor key are likely to be a hit before 2020. However, the true value for the coefficient for mode can also be negative according to @fig-model-result-ci since the interval passes through 0.
